Simple Trial Report

### 2020-12-07 inst/jss/supplemental-material/listdown-jss.rmd
Minor changes.

---
documentclass: jss
author:
  - name: Michael Kane
    affiliation: Yale University
    address:
    - 60 College Street
    - New Haven, CT 06510, USA
    email: michael.kane@yale.edu
  - name: Xun (Tony) Jiang
    affiliation: Amgen Inc.
    address:
    - One Amgen Center Drive
    - Thousand Oaks, CA 91320-1799, USA
    email: xunj@amgen.com
  - name: Simon Urbanek
    affiliation: The University of Aukland
    address:
    - 38 Princes Street
    - Auckland Central, Aukland 1010, NZ
    email: s.urbanek@auckland.ac.nz
title:
  formatted: "On the Programmatic Generation of Reproducible Documents"
  # If you use tex in the formatted title, also supply version without
  plain:     "On the Programmatic Generation of Reproducible Documents"
  # For running headers, if needed
  short:     "On the Programmatic Generation of Reproducible Documents"
abstract: >
  Reproducible document standards, like \proglang{R} Markdown, facilitate the programmatic creation of documents whose content is itself programmatically generated. While these documents are generally not complete in the sense that they will not include prose content, generated by an author to provide context, a narrative, etc., programmatic generation can provide substantial efficiencies for structuring and constructing documents. This paper explores the programmatic generation of reproducible by distinguishing components that can be created by computational means from those requiring human-generated prose, providing guidelines for the generation of these documents, and identifying a use case in clinical trial reporting. These concepts and use case are illustrated through the \pkg{listdown} package for the \proglang{R} programming environment, which is is currently available on the Comprehensive R Archive Network (CRAN).
keywords:
  # at least one keyword must be supplied
  formatted: [reproducibility, reproducible documents, data presentation]
  plain:     [reproducibility, reproducible documents, data presentation]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

@@

\proglang{R} Markdown [@baumer2014] facilitates the construction of computationally reproducible documents by allowing authors to insert \proglang{R} code for data processing, exploration, analysis, table-making, and visualization directly into structured, electronic documents. The resulting documents are made up of these chunks of \proglang{R} code, which we will refer to as _computational components_ since they are generated by computational means, as well as _narrative components_, which (in scientific writing) is prose intended to contextualize computational components, provide background, define goals, establish themes, and convey the results of the document.
The integration of narrative and computational components was identified as "Literate Programming" by @knuth1984 and software tools, like Sweave [@leisch2002], have supported this functionality for almost two decades.

\proglang{R} Markdown is particularly popular with its success likely being driven by two factors. The first is the relative ease with which these documents can be constructed. While \LaTeX \ is more expressive, it is relatively technical and requires an investment in time to become proficient. By contrast \proglang{R} Markdown documents are easier to create and format and, when the document is used to create \LaTeX, formatting can be passed through to the underlying .tex file. The second factor driving adoption is likely its support for creating _modifiable_ documents, namely Microsoft Word documents. Researchers and analysts, especially those creating applied statistical analyses, often collaborate with domain experts with less technical knowledge. In these cases, the analyst focuses on creating the computational components and narrative components related to results and interpretation. After this initial document is created, the domain expert is free to develop narrative components directly in the document without needing to go through the analyst.

Since computational components are, by definition, computationally derived objects and \proglang{R} Markdown is a well-defined standard, it is possible to programmatically create \proglang{R} Markdown documents with computational components, which is the focus of this paper. Generating documents in this manner has two appealing characteristics. First, it allows us to distinguish the presentation of analytical results from other steps in a data science or data processing pipeline. These other steps including cleaning and analysis often require their own environment and configuration with requirements very different than the computational needs of creating a presentation. By separating these components each can be developed independently. At the same time, by specifying a contract for the output of those objects, we can establish a consistent means by which processed data can be passed to systems for presenting those data in a structured way. The second reason for programmatic creation of R Markdown documents is convenience. In collaborative environments, especially in the early stages, large numbers of graphs and tables are generated and discussed. By collecting these artifacts and structuring them consistently, we can quickly iterate upon and restructure the resulting documents to more clearly present the data without needing to spend time on the creation of the presentation document.

@@

The \pkg{listdown} package itself is relatively simple with 6 distinct methods that can be easily incorporated into existing analysis pipelines for automatically creating documents that can be used for data exploration and reviewing analysis results as well as a starting point for a more formal write up. These methods include:

\begin{itemize}
\item{\bf as\_ld\_yml() }{- turn a computational component list into YAML with class information}
\item{\bf ld\_cc\_dendro() }{- create a dendrogram from a list of computational components}
\item{\bf ld\_chunk\_opts() }{- apply chunk options to a presentation object}
\item{\bf ld\_ioslides\_header() }{- create an ioslides presentation header}
\item{\bf ld\_make\_chunks() }{- write a listdown object to a string}
\item{\bf ld\_rmarkdown\_header() }{- create an R Markdown header}
\item{\bf ld\_workflowr\_header() }{- create a worflowr header}
\item{\bf listdown() }{- create a listdown object to create an \proglang{R} Markdown document}
\item{\bf print.listdown()}{- print the listdown options for \proglang{R} Markdown document creation}
\end{itemize}

The rest of this paper is structured as follows. The next section goes over basic usage and commentary. This section is meant to convey the basic approach used by the package and shows how to describe an output document using \pkg{listdown}, create a document, and change how the presentation of computational components can be specialized using \pkg{listdown} decorators. With the user accustomed to the package's basic usage, section 3 describes the design of the package. Section 4 goes over advanced usage of the package including adding initialization code to and outputted document as well as how to control chunk-level options. Section 5 provides a simplified case study of how the package is currently being used in clinical trial reporting. Section 6 concludes the paper with a few final remarks on the general types of applications where \pkg{listdown} has been shown effective.

# Basic usage and overview

@@

As a starting point, we will consider the a list of visualizations from the Anscombe data set below. The list is composed of four \pkg{ggplot2} [@wickham2016] elements (named Linear, Non Linear, Outlier Vertical, and Outlier Horizontal) each containing a scatter plot from the Anscombe Quartet - made available in the \pkg{datasets} package [@R] as shown in Figure \ref{cc1}. From the \code{computational_components} list, we would like to create a document with four sections with names corresponding to the list names, each containing their respective visualizations. The structure of a document derived from the \code{computational_components} list can be visualized using the \code{ld_cc_dendro()} function, and its output is below.

@@

The \code{ld} object, along with the computational components in the \code{comp-comp.rds} file are sufficient to to create the sections, subsections, and \proglang{R} chunks of a document. The only other thing requires to create the document is the header. The listdown package currently supports regular R Markdown and \pkg{workflowr} as \code{yml} objects from the \pkg{yaml} package [@yaml]. These objects are stored as named lists in \proglang{R} and are easily modified to accommodate document parameters. A complete document can then be written to the console using the code shown below. It could easily be written to a file for rendering using the \code{writeLines()} function, for example.

```{r eval = FALSE}
doc <- c(
  as.character(ld_rmarkdown_header("Anscombe's Quartet",
                                   author = "Francis Anscombe",
                                   date = "1973")),
  ld_make_chunks(ld))

writeLines("anscome-example.rmd")

doc
```

```{r eval = TRUE, echo = FALSE}
doc <- c(
  as.character(ld_rmarkdown_header("Anscombe's Quartet",
                                   author = "Francis Anscombe",
                                   date = "1973")),
  ld_make_chunks(ld))

doc
```

The \code{listdown()} function provides _document-wide_ R chunk options for displaying computational components. The chunk options are exactly the same as those in the R markdown document and can be used to tailor the default presentation for a variety of needs. The complete set of options can be found in the R Markdown Reference Guide [@rmarkdownref]. As a concrete example, the code used to create present the plots could be hidden in the output document using the following code.

@@

The first example is simple in part because the ggplot objects both contains the data we want to display and, at the same time, provides the mechanism for presenting them - rendering them in a graph. However, this is not always the case. The objects being stored in the list of computational components may not translate directly to the presentation in a document. In these cases, a function is needed that takes the list component and returns an object to be displayed. For example, suppose that, along with showing graphs from the Anscombe Quartet, we would like to include the data themselves. We could add the data to the \code{computational_components} list and then create the document with:

@@

List names in the \code{decorator} argument provide a key to which a function or method is mapped. The underlying decorator resolution is implemented for a given computational component by going through decorator names sequentially to see if the component inherits from the name using the \code{inherits()} function. The function or method is selected from the corresponding name which the element first inherits from. This means that when customizing the presentation of objects that inherit from a common class, the more abstract classes should appear at the end of the list. This will ensure that specialized classes will be encountered first in the resolution process. It should be noted that an object's type is first checked against the decorator name list and then checked to see if it is a list. This allows a user to both decorate a list and retain \code{"list"} in its class attributes.

@@

@@

## Initialization code

The \code{listdown()} function facilitates the insertion of initialization code through the \code{init_expr} argument. Code is inserted immediately after the libraries are loaded in the R Markdown document. In general, it is suggested that the number of initial expressions be kept small so that the \proglang{R} Markdown document is easy to read. If a large number of functions are required by the target \proglang{R} Markdown document then they can be put into a file and sourced using the initial expression. As an example, suppose we are creating an html document and presenting data using the \code{datatable()} function. However, we do not want to include the search capabilities provided. This can be easily accomplished by creating a new function, \code{datatable_no_search()}, created using the \code{partial()} function \cite{purrr} to partially apply \code{list(dom = 't')} to the \code{options} argument of \code{datatable}.

```{r}
saveRDS(comp_comp2, "comp-comp2.rds")
ld <- listdown(load_cc_expr = readRDS("comp-comp2.rds"),
               package = c("ggplot2", "DT", "purrr"),
               decorator = list(ggplot = identity,
                                data.frame = datatable_no_search),
               init_expr = {
                 datatable_no_search <- partial(datatable,
                                                options = list(dom = 't'))
                 },
               echo = FALSE)

ld_make_chunks(ld)[2:10]
```

@@

@@

make_surv_cc <- function(trial, treat, surv_cond_chars) {
  table_1 <- trial %>%
    tbl_summary(by = all_of(treat)) %>%
    gtsummary::as_flextable()

>>>>

---
documentclass: jss
author:
  - name: Michael Kane
    affiliation: Yale University
    address:
    - 60 College Street
    - New Haven, CT 06510, USA
    email: michael.kane@yale.edu
  - name: Xun (Tony) Jiang
    affiliation: Amgen Inc.
    address:
    - One Amgen Center Drive
    - Thousand Oaks, CA 91320-1799, USA
    email: xunj@amgen.com
  - name: Simon Urbanek
    affiliation: The University of Aukland
    address:
    - 38 Princes Street
    - Auckland Central, Auckland 1010, NZ
    email: s.urbanek@auckland.ac.nz
title:
  formatted: "On the Programmatic Generation of Reproducible Documents"
  # If you use tex in the formatted title, also supply version without
  plain:     "On the Programmatic Generation of Reproducible Documents"
  # For running headers, if needed
  short:     "On the Programmatic Generation of Reproducible Documents"
abstract: >
  Reproducible document standards, like \proglang{R} Markdown, facilitate the programmatic creation of documents whose content is itself programmatically generated. While programmatic content alone may not be sufficient for a rendered document since it does not include prose (content generated by an author to provide context, a narrative, etc.) programmatic generation can provide substantial efficiencies for structuring and constructing documents. This paper explores the programmatic generation of reproducible documents by distinguishing components that can be created by computational means from those requiring human-generation, providing guidelines for the generation of these documents, and identifying a use case in clinical trial reporting. These concepts and use case are illustrated through the \pkg{listdown} package for the \proglang{R} programming environment, which is is currently available on the Comprehensive R Archive Network (CRAN).
keywords:
  # at least one keyword must be supplied
  formatted: [reproducibility, reproducible documents, data presentation]
  plain:     [reproducibility, reproducible documents, data presentation]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

@@

\proglang{R} Markdown [@baumer2014, @rmarkdownbook] facilitates the construction of computationally reproducible documents by allowing authors to insert \proglang{R} code for data processing, exploration, analysis, table-making, and visualization directly into structured, electronic documents. The resulting documents are made up of these chunks of \proglang{R} code, which we will refer to as _computational components_ since they are generated by computational means, as well as _narrative components_, which (in scientific writing) is prose intended to contextualize computational components, provide background, define goals, establish themes, and convey results. These documents are then used to render output documents, for users to read in the form of .html, .pdf, .doc, or other formats using the \pkg{knitr} package [@knitrbook].

The integration of narrative and computational components was originally identified as "Literate Programming" by @knuth1984 and software tools, like Sweave [@leisch2002], have supported this functionality for almost two decades. However, more recently, \proglang{R} Markdown has become particularly popular with its success likely being driven by two factors. The first is the relative ease with which these documents can be constructed. While \LaTeX \ is more expressive, it is relatively technical and requires an investment in time to become proficient. By contrast \proglang{R} Markdown documents are easier to create and format and, when the document is used to create \LaTeX, formatting can be passed through to the underlying .tex file. The second factor driving adoption is likely its support for creating _modifiable_ documents, namely Microsoft Word documents. Researchers and analysts, especially those creating applied statistical analyses, often collaborate with domain experts with less technical knowledge. In these cases, the analyst focuses on creating the computational components and narrative components related to results and interpretation. After this initial document is created, the domain expert is free to develop narrative components directly in the document without needing to go through the analyst.

Since computational components are, by definition, computationally derived objects and \proglang{R} Markdown is a well-defined standard, it is possible to programmatically create \proglang{R} Markdown documents with computational components, which is the focus of this paper. Generating documents in this manner has two appealing characteristics. First, it allows us to distinguish the presentation of analytical results from other steps in a data science or data processing pipeline. Other steps including cleaning and analysis often require their own environment and configuration with requirements very different than the computational needs of creating a presentation. By separating these components each can be developed independently. At the same time, by specifying a contract for the output of those objects, we can establish a consistent means by which processed data can be passed to systems for presenting those data in a structured way. The second reason for programmatic creation of R Markdown documents is convenience. In collaborative environments, especially in the early stages, large numbers of graphs and tables are generated and discussed. By collecting these artifacts and structuring them consistently, we can quickly iterate upon and restructure the resulting documents to more clearly present the data without needing to spend time on the creation of the presentation document.

@@

The \pkg{listdown} package itself is relatively simple with 10 distinct methods that can be easily incorporated into existing analysis pipelines for automatically creating documents that can be used for data exploration and reviewing analysis results as well as a starting point for a more formal write up. These methods include:

\begin{itemize}
\item{\bf as\_ld\_yml() }{- turn a computational component list into YAML with class information}
\item{\bf ld\_cc\_dendro() }{- create a dendrogram from a list of computational components}
\item{\bf ld\_chunk\_opts() }{- apply chunk options to a presentation object}
\item{\bf ld\_ioslides\_header() }{- create an ioslides presentation header}
\item{\bf ld\_make\_chunks() }{- write a listdown object to a string}
\item{\bf ld\_rmarkdown\_header() }{- create an R Markdown header}
\item{\bf ld\_workflowr\_header() }{- create a worflowr header}
\item{\bf ld\_write\_file() }{- write to and R Markdown file}
\item{\bf listdown() }{- create a listdown object to create an \proglang{R} Markdown document}
\item{\bf print.listdown()}{- print the listdown options for \proglang{R} Markdown document creation}
\end{itemize}

The rest of this paper is structured as follows. The next section goes over basic usage and commentary. This section is meant to convey the approach used by the package and shows how to describe an output document using \pkg{listdown}, create a document, and change how the presentation of computational components can be specialized using \pkg{listdown} decorators. With the user accustomed to the package's basic usage, section 3 describes the design of the package. Section 4 goes over advanced usage of the package including adding initialization code to and outputted document as well as how to control chunk-level options. Section 5 provides a simplified case study of how the package is currently being used in clinical trial reporting. Section 6 concludes the paper with a few final remarks on the general types of applications where \pkg{listdown} has been shown effective.

# Basic usage

@@

As a starting point, we will consider the a list of visualizations from the Anscombe data set below. The list is composed of four \pkg{ggplot2} [@wickham2016] elements (named Linear, Non Linear, Outlier Vertical, and Outlier Horizontal) each containing a scatter plot from the Anscombe Quartet - made available in the \pkg{datasets} package [@R]. From the \code{computational_components} list, we would like to create a document with four sections with names corresponding to the list names, each containing their respective visualizations. The structure of a document derived from the \code{computational_components} list can be visualized using the \code{ld_cc_dendro()} function, and its output is below.

@@

The \code{ld} object, along with the computational components in the \code{comp-comp.rds} file are sufficient to to create the sections, subsections, and \proglang{R} chunks of a document. The only other thing required to create the document is the header. The \pkg{listdown} package currently supports regular R Markdown and \pkg{workflowr} as \code{yml} objects from the \pkg{yaml} package [@yaml]. These objects are stored as named lists in \proglang{R} and are easily modified to accommodate document parameters. A complete document can then be written to the console using the code shown below. It could easily be written to a file for rendering using the \code{ld_write_file()} function, for example.

```{r eval = FALSE}
ld_write_file(ld_rmarkdown_header("Anscombe's Quartet",
                                  author = "Francis Anscombe",
                                  date = "1973"),
              ld,
              "anscome-example.rmd")
```

The \code{listdown()} function provides _document-wide_ R chunk options for displaying computational components. The chunk options are exactly the same as those in the R Markdown document and can be used to tailor the default presentation for a variety of needs. The complete set of options can be found in the R Markdown Reference Guide [@rmarkdownref]. As a concrete example, the code used to create present the plots could be hidden in the output document using the following code.

@@

The first example is simple in part because the \code{ggplot} objects both contains the data we want to display and, at the same time, provides the mechanism for presenting them - rendering them in a visualization However, this is not always the case. The objects being stored in the list of computational components may not translate directly to the presentation in a document. In these cases, a function is needed that takes the list component and returns an object to be displayed. For example, suppose that, along with showing graphs from the Anscombe Quartet, we would like to include the data themselves. We could add the data to the \code{computational_components} list and then create the document with:

@@

List names in the \code{decorator} argument provide a key to which a function or method is mapped. The underlying decorator resolution is implemented for a given computational component by going through decorator names sequentially to see if the component inherits from the name using the \code{inherits()} function. The function or method is selected from the corresponding name which the element first inherits from. This means that when customizing the presentation of objects that inherit from a common class, the more abstract classes should appear at the right-end of the list. This will ensure that specialized classes will be encountered first in the resolution process. It should be noted that an object's type is first checked against the decorator name list and then checked to see if it is a list. This allows a user to both decorate a list and retain \code{"list"} in its class attributes.

@@

It should be noted that it is not possible to decorate a list and an attempt to do so results in an error. This is because, when generating a document, the `ld_make_chunks()` function recursively descends into the list of computational components. The decision to descend is made based on the type of the element being visited. If it is a list, then it descends otherwise it presents the object. However, a list can't designate both an object to hold elements for presentation and an object to be presented. To present arbitrary list elements, including lists, a user may add class information to a list element and a corresponding decorator.

@@

## Setup and initialization code

The \code{listdown()} function facilitates the insertion of setup and initialization code through the \code{setup_expr} and \code{init_expr} arguments. If and argument is provided to the \code{setup_expr}, then the first code chunk of the document will have the specified code inserted. This code chunk is named "setup" and the include parameter is set to \code{FALSE}. When the \code{init_expr} argument is specified,
code is inserted immediately after the libraries are loaded in the R Markdown document. In general, it is suggested that the number of initial expressions be kept small so that the \proglang{R} Markdown document is easy to read. If a large number of functions are required by the target \proglang{R} Markdown document then they can be put into a file and sourced using the initial expression. As an example, suppose we are creating an html document and presenting data using the \code{datatable()} function. However, we do not want to include the search capabilities provided. This can be easily accomplished by creating a new function, \code{datatable_no_search()}, created using the \code{partial()} function \cite{purrr} to partially apply \code{list(dom = 't')} to the \code{options} argument of \code{datatable}.

```{r}
saveRDS(comp_comp2, "comp-comp2.rds")
ld <- listdown(load_cc_expr = readRDS("comp-comp2.rds"),
               package = c("ggplot2", "DT", "purrr"),
               decorator = list(ggplot = identity,
                                data.frame = datatable_no_search),
               setup_expr = knitr::opts_chunk$set(echo = FALSE),
               init_expr = {
                 datatable_no_search <- partial(datatable,
                                                options = list(dom = 't'))
                 })

ld_make_chunks(ld)[2:14]
```

@@

## Presenting base graphics

Since a base plot does not encapsulate the state and ability to present the plot, like a `ggplot` object, for example it is not possible to assign a base graphic to a list element and present it. Instead, it is recommended that the corresponding data is held as a list element, with base graphic options, and that that element is given a class and a corresponding decorator. Using this approach the decorator generates the visualization though a call to base graphics when the output document is rendered.

@@

make_surv_cc <- function(trial, treat, surv_cond_chars) {
  table_1 <- trial %>%
    tbl_summary(by = all_of(treat)) %>%
    gtsummary::as_flex_table()
